\section{Second Day of Easter: Faith or Hardening}

\begin{quote}

Matthew 28:9–15: And as they went to tell his disciples, behold, Jesus met them and said, “Hail.” And they came and took hold of his feet and worshiped him. Then Jesus said to them, “Do not fear. Go, tell my brothers that they are to go into Galilee, and there they shall see me.”

And while they were going, behold, some of the guard came into the city and told the chief priests all that had happened. And they assembled with the elders and took counsel, and they gave the soldiers many silver coins and said, “Say: His disciples came by night and stole him while we slept. And if this comes to the governor’s ears, we will persuade him and keep you free from trouble.” And they took the silver coins and did as they were instructed. And this saying has been spread among the Jews unto this day.

\end{quote}

\bigskip

When Jesus was born, an angelic message concerning it sounded to the shepherds; they believed the message, went to Bethlehem, found the Child and his mother, and praised Israel’s God, who had given his people a Savior, the Lord Christ in David’s city. To Gentile wise men the star brought tidings of the birth of the King of the Jews, and they believed the message, and they came and worshiped the Child in Mary’s arms. But to Herod the Great, the king in Jerusalem, that same message came through the wise men from the East, and he believed, in the same manner as the devils believe that God is one, and he trembled for fear, and he hardened his heart and sought to slay the Child.

So it goes also with Jesus’ resurrection from the dead. Our text today tells of it, for our encouragement and warning, how it becomes for some a fragrance of life unto life, but for others a fragrance of death unto death. And that question demands its answer also from us: What do we do with Jesus’ resurrection from the dead? And where do we stand with the risen Jesus? Upon the answer depends our salvation or our perdition, eternal life or eternal death. May God’s Spirit, who alone can give the right answer within believing hearts, help us that we may believe and live.

Jesus Christ’s resurrection is a miracle no one can honestly deny; that is to say, it is impossible to deny that Jesus is risen without lying. All who deny Jesus’ resurrection know that they deny that which is true. Because they tremble at the thought of this miracle, they prefer to say that they do not believe it.

Jesus Christ’s resurrection is a miracle no one can honestly deny, and that means it powerfully proves that he is God’s Son. Therefore it also brings life and blessedness for those who believe in the risen One and worship him.

Faith or hardening is therefore the real choice that lies before us. The dull, indifferent sleep in which so many of the so‑called Christians find themselves is in reality already a sure sign of death, and there is no salvation for any of these sleeping ones unless they give heed to the Word which sounds to them also out from the open grave: Awake, you who sleep, and arise from the dead.

The women of whom our text speaks are for us an example of those who believe unto the salvation of the soul; the chief priests, on the other hand, are of those who shrink back unto their destruction.

The women had been at the open grave, and an angel had spoken to them a blessed word: “Do not fear; for I know that you seek Jesus, the crucified. He is not here; for he is risen, as he said. Come here, see the place where the Lord lay. And go quickly and tell his disciples that he is risen from the dead. And behold, he goes before you into Galilee; there you shall see him. Behold, I have told you.”

The women believed the Word and obeyed the command. With fear and great joy they ran to bring the glad message to the eleven. They had never been so shaken; their hearts trembled with fear and joy, and they hurried off, that all who loved Jesus might hear what they had seen and heard.

Blessed are they who believed the message they heard; for it went with them as with Nathanael: they received to see greater things. Jesus himself met those who believed the word from the angel’s mouth, and he greeted them with the glad word: “Hail!” It was the first word from the lips of the Risen One to those for whose sake he had gone into death; it bore with it to those who heard it a joyful assurance of his divine majesty, of his infinite love, and of his perfect victory over sin and death and all the power of evil. The Crucified had conquered, the Dead was alive, God’s glory was revealed among men as never before.

The women fell to their knees before the Risen One; it was the Son of God who stood before them, and he did not refuse to receive worship from them. So shall his believing congregation worship him and exalt and praise his Name. He is not only the one through whom all was created and through whom all creation has its natural life; he is also the firstborn from the dead, through whom eternal life and incorruptibility are brought to light; he is the Head of the body, the congregation, and what the old creation lost through sin, that the new creation receives back in the risen Savior. Bow before him, my soul, and worship; for the Lamb who was slain is worthy to receive honor and praise and power unto eternity.

It was a blessed hour for the women; such is only the hour in which the longing soul finds its Savior and a poor sinner receives the forgiveness of sins. It is the holy joy of faith which the world has never tasted and never shall taste. All perishable joys together weigh nothing at all against the soul’s infinite blessedness in Jesus’ fellowship and worship. Do you know this blessedness, brother? It is the fruit of faith in the risen Jesus. 

For if he is risen, then sin is atoned and God’s wrath taken away. If he is risen, then that life is won which death cannot destroy. If he is risen, then our faith is not in vain, but mighty to raise us with Christ, make us alive with him, glorify us with him, and set us with him in the heavenly places, so that in the midst of death’s and corruption’s world, while we ourselves still bear the body of death, yet we are God’s children, and Christ lives in us, so that we are dead, and our life is hidden with Christ in God. Blessed and praised be he who was dead, and behold, he lives unto all eternity.

And when the women had seen Jesus and were anew assured of his resurrection, Jesus sends them to do his work and bear his Word to the apostles. All who have found life in faith in God’s Son must labor in his service. No one can love Jesus in an evil and unbelieving world without confessing him. Therefore he lets his believing ones remain in the body and live upon the earth, that they shall be his witnesses. And here Jesus sends the women to the apostles with the loving greeting: “Tell my brothers that they shall go into Galilee; there they shall see me.” In the hour of his resurrection Jesus calls the disciples brothers. That too is the effect and fruit of the resurrection. Now they shall not only hear his Word and learn from his wisdom; now they shall also become partakers of his divine life, which is offered and given for them, and which by the Spirit shall be poured out into them.

That is the congregation. It is possible only through Jesus’ death and resurrection. Only by the power of the atonement can sinners draw near to the Lord in his sanctuary; only by the resurrection from the dead can Jesus’ immortal life be communicated to those who were dead in sins and trespasses, and who were reconciled to God through his Son’s death.

Blessed are those who truly believe in the Risen One and are born again to a living hope through Jesus Christ’s resurrection from the dead.

But those who know Jesus only in order to hate him — what can they do with Jesus’ resurrection? All the sleeping souls who love to sleep, and who above all things hate the light from Christ’s open grave and Easter morning’s awakening earthquake and its signs and wonders — what must they do to stay asleep?

There is one thing they must do, if the risen Savior shall not awaken and save and make them alive; they must harden themselves like the chief priests and deny for themselves and others the living truth, concerning which they know full well cannot be denied.

When the soldiers came and told the chief priests what they had seen and heard, then the chief priests understood very well that it had happened as Jesus had said: “After three days I rise again.” But they would not that it should be true; for if it were true, then they must acknowledge their guilt and condemn themselves and all their conduct, when they had denied that Jesus was the Christ and had slain the Prince of life. It should not be true; for then they themselves were wretched criminals, but Jesus’ disciples were those who had understood God’s counsel and chosen the way of salvation and life. They would not repent; therefore they must harden themselves. And since the soldiers were witnesses of their shame, they did with them as with Judas; they gave them many silver coins, that they should become traitors against the truth and spread falsehood instead. Like so many other rulers in state and church, these too had learned to believe in the power of money. They thought that before the power of money even the truth, God’s living truth, should lie subdued. And how did that turn out? Truly, the soldiers at the grave and the seal upon the stone and the very money with which the false testimony was bought have become speaking, yes crying witnesses of Jesus’ resurrection. For why should the frightened guard receive money, if it did not know more than the chief priests wished to have spread among the people?

Yet even today it is possible for men to close heart and mind to the Gospel of Christ’s resurrection from the dead. Even now men in the midst of Christendom can cling to earthly things with such passion and desire that they would rather reject Christ and the hope of the resurrection and eternal life than repent and believe in the Crucified. The sleeping members of the congregation, who are awakened neither by the earthquake nor by the angel’s voice, often resemble the chief priests who paid money that the truth should not become known and condemn them. Or are there not many carnal men who would rather pay money than repent? So long as they are not disturbed in their carnal security and false peace, they will sacrifice much — to their own destruction.

But if Jesus is risen from the dead, then it becomes a matter of life or death what you do with it. and with the Gospel concerning it. Jesus died for your sins and rose for your justification — does that mean nothing to you? He opened heaven for you; are you then so busy with the earth and earthly things that you care nothing for it? If you live as though there were no resurrection, and all your conduct is only a testimony that you seek the earthly things, then you do not believe God’s blessed Gospel, and with every day that passes you harden your heart more and more against the truth.

If Jesus is risen, then let us turn to him in whom there is deliverance from death, and being raised with him seek those things which are above, where Christ sits at God’s right hand. For our citizenship is in the heavens, from where also we await as Savior the Lord Jesus Christ, who shall transform our body of humiliation to be conformed to his body of glory, according to the power whereby he is able also to subject all things unto himself. 

Amen.

