
% Update 01 Mar 2026: Locked the opening bible verse to DAN 1819 with intentional sin trolovede Hustru (his betrothed wife). Eliminated almost all KJV influence, and set emotional volatility.
 
\section{First Day of Christmas: A Great Joy}


\begin{quote}
Luke 2:1–14. In those days a decree went out from Caesar Augustus that the whole world should be registered. This first registration took place when Quirinius was ruler in Syria. All went to be registered, each to his own city.

Joseph also went up from Galilee, from the city of Nazareth, to Judea, to the city of David which is called Bethlehem, because he was of the house and lineage of David, to be registered together with Mary, his betrothed wife, who was with child.

While they were there, the time came for her to give birth. She gave birth to her firstborn son, and wrapped him in swaddling cloths and laid him in a manger, because there was no room for them in the inn.

There were shepherds in that same region, staying out in the fields and keeping watch by night over their flock. An angel of the Lord stood before them, and the glory of the Lord shone around them, and they were terrified. And the angel said to them, “Do not fear, for I proclaim to you a great joy which shall befall all people. For to you is born this day a Savior, who is the Lord Christ, in the city of David. And this shall be a sign to you: you will find a child wrapped in swaddling cloths, lying in a manger.”

Suddenly the angel was joined by a multitude of the heavenly host, praising God and saying, “Glory to God in the highest! and on earth peace! and good will among people!”
\end{quote}
\bigskip


It is written of the shepherds on the fields of Bethlehem that “they were greatly afraid” when the angel of the Lord stood before them and the glory of the Lord shone around them.

Why were they afraid? And do you also fear, O heart, when the light of the Lord shines around you? What is it that you fear?

He who has experienced it knows well what it is that makes a human heart so fearful and anxious when the light of the Lord breaks in and it becomes bright as day within. And many, many who have not allowed the light of the Lord to shine into the heart know very well why they do not dare open the door of the heart to the light and the day.

Oh, how carefully they strive to keep it dark in the closed chamber of the heart, so that it may not be revealed how things truly look within. They sense how badly things stand; they keep it dark so that naked reality may not laugh them in the face. They do not want to see it themselves, and when it is too dark for their own eye, they so easily deceive themselves with the false comfort: No one sees it; God does not see it!

But “the Lord sees,” it is written.

Foolish person, open the door of the heart, that the Lord's light may break in. It is better that you be cast down in fear and terror when the light comes in, than that you try in vain to shut it out.

Futile? Yes, futile; for even if it should succeed for a short while to shut out the light of grace, it will be utterly futile when the light becomes a consuming fire of wrath.

O, if you tremble before the glory of the Lord, my poor heart—open up, open up! If you think with a half-suppressed sigh of Christmas as it comes, of the Gospel as it sounds, of the child in the manger, of Mary in the stable, of the shepherds in the fields—if you think: If only I were a child again, that I might rejoice as before—open up, open up to the light of the Lord!

Your heart is in trouble, but open up, open up to the light of the Lord, and there is still hope.

For what is it that you lack? It is the coldness of the world, it is the darkness of doubt, it is the burning fog of reason, it is the misery of sin that fills the heart. They are enemies of your joy and peace and blessedness that have pressed in and settled inside the heart. Let only the light of the Lord shine in, so that the nature of the enemies may be made known, and you may see the corruption they have wrought.

\textbf{And when you then in terror must cry out: my sin, my sin! then comes the angelic message: Do not fear! The Savior is here! A great joy shall befall all the people — and to you this day}

A great joy — yes, a joy from heaven itself, the joy of children, the jubilation of angels

For a Savior is born to you; “He will save his people from their sins.” For this reason he has come into the world; for this reason he comes to you.

If your old sins weigh upon you, he will forgive them. If you think you are so unclean and filthy and ragged and torn that you dare not come before God—oh see, it is but a little child who stretches out his arms toward you and smiles at you; he does not take account of what you have been, where you have gone, how wretched and miserable you have become, if only you will now receive the little child and give him room with you.

Then you shall experience joy, joy as never before. For if you took the child Jesus into heart and home, then it was God’s life and love you gave room within you. There came new life, there came new joy. You experienced that all the old has passed away and all has become new.

Is that the great joy? Yes, it is the only pure and perfect joy that a human heart can experience here on earth. When all the old, all sin, all sorrow, all distress is sunk into the depths of eternal mercy; when all old anxious memories are taken away; when all uncleanness is washed away in the blood of Jesus; when you are permitted to begin life anew without any oppressive guilt, without any constricting fear, without any gnawing resentment, and you may live saved and free in the bright day of grace.

Then there is joy. Heavenly joy in a human soul.

But there are few who taste this joy fully and completely. And yet it is precisely therein that the glory of the Christian faith proves itself in all its fullness.

It seems so small to the world to go to the manger in Bethlehem; it seems so contemptible in the eyes of the world to become a child again; it seems so impossible for one who is old to be born anew; and yet it is precisely this that is the glad message of Christmas: “A son is born to us, a child is given to us.”

Our reason says: Let us become better people, let us become more rational, wiser, more calculating, more cautious people.

Our conscience, inflamed by God’s law, says: Become righteous, become holy, become good!

But God’s Gospel says: Come to the manger of Jesus and become a child again!

That is the great joy. That is the new thing which God has done, and which fills our mouth with jubilation and our tongue with laughter: new life, eternal life, imperishable life given by God in his own Son, given to everyone who believes in the Son.

It seems so offensive to our old fleshly reason. And yet, if you are to receive an eternal life, does it make any sense to try to transform your old, perishable life into eternal life by your works?

Can an old garment become new by having ever so many patches sewn onto it?

Can perishability ever become anything other than perishability, no matter how greatly you exert yourself?

But either you must receive an eternal life as a free, undeserved gift from God, or you can never, never attain it.

And here the gift is given; here is the true God and eternal life. Come and see! Come and take the child into your heart in living faith, and you yourself shall be born anew.

Is that possible? Ah no, for a human being this is impossible; but remember what was said of the child Jesus: For God nothing is impossible.

If you believe this miracle, then the miracle has already taken place in your own heart. Then you yourself are a child of God, born anew by the miraculous power of God.

That is true and living Christianity.

Come, let us go with the shepherds and see this thing that has come to pass, which the Lord has made known to us, and there shall be great joy in our hearts.

Joy is so greatly lacking in our Christianity because simple faith in Christ is so greatly lacking. Oh, that it might become better with us all at this Christmas feast! Let us then open the doors of our hearts wide and let the light of the Lord shine in; and when we are terrified over all our sin, all our self-love, all our unbelief, all our pride, all our self-righteousness, then let us also take the child Jesus in, and his holy love shall blot out sin and kindle the new heavenly love within, and we shall receive

a great joy.
