
\section{Twenty-Sixth Sunday after Trinity: The Judgment and the Life}

\begin{quote}

John 5:22–29. For neither does the Father judge anyone, but has given all judgment to the Son, that all should honor the Son, just as they honor the Father. Whoever does not honor the Son does not honor the Father who has sent him. Truly, truly, I say to you: Whoever hears my word and believes him who has sent me has eternal life and does not come into judgment, but has passed from death to life. Truly, truly, I say to you: The hour comes, and now is, when the dead shall hear the voice of the Son of God, and those who hear it shall live. For just as the Father has life in himself, so he has also given to the Son to have life in himself; and he has given him authority also to execute judgment, because he is the Son of Man. Do not marvel at this! For the hour comes when all who are in the graves shall hear his voice, and they shall come forth: those who have done good, to the resurrection of life; but those who have done evil, to the resurrection of judgment.

\end{quote}

\bigskip


The Father has given the Son all judgment; but he has also given him all life; he has given him to have life in himself. For the Father loves the Son and has given all things into his hand.

Thus equipped he comes before us, he who was so deeply abased and hung upon the Tree of the Curse. With such power has the Father crowned him whose head bled from the thorns of mockery.

In one hand he holds life, in the other he holds death. With one hand he offers restoration, with the other condemnation. Today he offers salvation; tomorrow he holds judgment.

In him the fullness of the Godhead is revealed bodily. In him, a man, the holy God comes before men with his omnipotence and grace and says: Honor the Son! No one can honor the Father without honoring the Son. Around him the armies of heaven join together and cry with a loud voice: “The Lamb who was slain is worthy to receive power and riches and wisdom and strength and praise and honor and blessing.”

The Son of Man has been given all power in heaven and on earth, and he is truly not far from any one of us.

Friend, have you listened to his voice, or have you had the courage to shut your ear in contempt when he spoke to you and said: “Here is life, there is judgment”?

O, consider yourself once more; here he comes forth before you again, behold him! Behold the Lamb of God, who bears the sin of the world! Far off in the desert’s desolation and loneliness he followed after you; your soul was drained and languishing; bitterly had the world mocked you with the empty cups of its lusts. Here Jesus stands before you, God’s own Son, himself full of wounds and pains; he stoops beneath you for your sake: “My child, come, and I will carry you upon my lacerated shoulders back to the Father’s house.” Friend, what did you answer?

He sought you as a mother seeks her child and found you seduced, enticed, and fallen among robbers; ah, how many an earthly friend, to whom you had looked up with expectation and trust, did not pass you by unfeeling in your misery! He came, whom you had forgotten, and who yet had suffered even hell’s torments for you; he came, he bent his thorn-crowned head over you and spoke lovingly to you; he healed your wounds and set you gently upon his own beast in order to bring you to safe shelter and care. Friend, what did you answer?

You writhed in anguish and pain; like sea-waves your sins went over your head, and hell’s abyss gaped toward you. “Is there no rescue for my poor soul?” you cried in despair. Then the Savior stood there again and said: “Come to me, you who labor and are burdened, and I will give rest to your soul.” Did you come? Could he draw you to Golgotha? Could he lift your eye to the cross, as Israel’s dying people lifted their gaze to the bronze serpent? Did he get to speak to your heart as to Jerusalem and say that your warfare is ended, your iniquity is atoned for; I have paid with my blood? Could he whisper to you this blessed word: “Take courage, your sins are forgiven you”? Could he draw you in faith to his breast in faith and give you eternal life according to his own words: “Whoever believes in the Son has eternal life”?

Friend, let us speak honestly with one another; how does it stand? Thousands of times the Son of God has stretched out his hands full of life and grace and offered you all this without money and without price. Ah, how have you received him, how have you answered him who has power to save completely all those who come to the Father through him?

It is drawing toward evening, and the night soon comes. Gloom and darkness and judgment are in the Lord’s right hand; his wrath is a consuming fire. He is not mocked, and whatever a man sows, that shall he also reap.

If you have had the courage or the levity to turn back his gracious offer of life and salvation, have you strength as well to hear his judgment? For there comes a day when all shall hear the voice of the Son of God; the dead shall come forth before his throne, and no one shall be judged without being heard.

“Is there anything that I could have done to save you, and which I neglected?” says the Son of Man. “In the desert, among robbers, in your soul’s anguish I visited you, and you paid no heed to me. With my Spirit and my word I knocked again and again at your heart’s door: ‘Let me in, that I may save you!’ but you would not. I became a man like you, despised for your sake and full of torment, and you would not acknowledge me.

I pointed to your sin and to hell’s open abyss, I showed you the way of life and death, I implored and begged you to return as a father calls a straying child home, I wept and said: ‘O, if only you knew, even today, what serves for your peace; O, if only I might hide you under my wings like a chick,’ — but you would not.

Yet I did not grow weary. I led you to Pilate; a man stands before you, beaten and bleeding: ‘Behold the Man!’ You turned away in indifference.”

At the foot of the cross you saw the Savior crowned with thorns and nailed fast, in unspeakable suffering and hellish anguish, all of it in order to save you. You were arrested for a moment; but then you turned again to the world’s levity and pleasure and cried: “Crucify! his blood be upon me and upon my children!”

Now he comes, the same Son of Man whom you have despised and rejected; the Lord has clothed him with power and majesty, and with the angels of heaven he comes in the clouds to enter into judgment with you and hold judgment. He who has life because he willed to become man, and has stretched it out to you, he has also received authority to hold judgment because he is the Son of Man. Have you anything to answer? Have you anything with which to excuse yourself?

Do you see him now? His countenance is changed; the bright, kindly look with which he drew you a thousand times has become a consuming fire. You tremble, and in unspeakable anguish you cast your eyes down. It was not he who rejected you; but you rejected him. Now you must receive the judgment.

“I was hungry and thirsty and naked and sick and a stranger and in prison; you neither gave me food nor drink, you neither clothed nor tended me, you neither came to me nor received me; depart from me, you cursed, into the eternal fire prepared for the devil and his angels!”

O, how dreadful, when you must learn that there is no peace for the ungodly, that he is tormented by a worm that never dies and a fire that is never quenched, and that in the midst of that tormenting exclusion from God’s blessed fellowship you shall yet have to acknowledge that Jesus Christ, the Son of Man, whom you despised and rejected, is Lord, to the glory of God the Father; he offered you grace and salvation, you chose judgment and torment; he offered you eternal life, and you chose an eternal death!
