
%FIXME: Check for "reduced emotional volatility." from the beginning of the text.

\section{First Day of Pentecost: The Spirit of Truth}

\begin{quote}

John 14:15–21. If you love me, keep my commandments. And I will pray to the Father, and he will give you another Advocate so that he may be with you forever, the Spirit of truth, whom the world cannot receive, for it neither sees him nor knows him; but you know him, for he abides with you and will be in you. I will not leave you as orphans; I come to you. Yet a little while, and the world sees me no more, but you see me; for I live, and you will live. On that day you will know that I am in my Father, and you in me, and I in you. Whoever has my commandments and keeps them, that is the one who loves me. And whoever loves me will be loved by my Father, and I will love him and reveal myself to him.

\end{quote}

\bigskip


Jesus is the truth and the life; therefore the Spirit whom he sends is the Spirit of truth and the Spirit of life.

A human life without Jesus and the Spirit of Jesus is a life without God; for Jesus alone is the mediator between God and men; therefore it is only through Jesus that we can receive life and Spirit from God.

Therefore a human life without Jesus is a life without truth, a life of falsehood. You seem to live, yet you do not live in truth. Only the appearance remains of that life which God gave the man he had created. Instead of fellowship with the eternal fountain of life, the heavenly Father, there has come surrender to the world with its vain desire, which flares up and quickly burns out like a fire in thorns. What is this but deception and falsehood, when man seeks to satisfy the hunger and thirst of his immortal soul with the perishable things of the world? It is more offensive than the prodigal son who filled his belly with swine’s food.

In falsehood lives the one who sinks an immortal soul into the perishableness of the world. Therefore he wins only death. In a little while he is separated from the world and the world from him, and eternal hunger and thirst become the lot of the soul, if through endless ages it has nothing but the consuming longing after the perishable and sensual things which can no longer be had.

Therefore, if you love the world, you deceive yourself; for the world passes away with its desire; and when the world is gone, what then have you to love? You are left with nothing — nothing but eternal torment and eternal death.

But Jesus is the truth and the life. And as the soul is united with the world when it loves the world, so also love for Jesus is the way to take part in his truth and his life.

Love Jesus! 

You know very well how this happens. For he has loved you first and has given his life for you upon the cross. He has stretched out his arms and hands toward you and said: come to me, you poor, sinful man, I will give you forgiveness of sins and rest for your soul. If you have cast yourself into his arms and found peace with God, then you have learned to love him who drew you up out of perdition and gave you eternal salvation.

Love Jesus! 

It consists in keeping his commandments; and his commandments are not burdensome, for it is this: to believe in his name and love one another, as he commanded us.

If this love for Jesus is in you, then you are partaker of his Spirit, who is the Spirit of truth and the Spirit of life. Whoever loves the world receives the Spirit of the world. Whoever loves Jesus receives the Spirit of Jesus. It is the mighty Spirit of truth who tears away from our soul the deceit of sin and shows us the abyss of perdition which hides itself behind the smiling mask of sinful desire. It is the Spirit of truth who lets the soul see the terror of sin, yet also points to him who is the Savior of sins, and who shows us all the way to the Father’s heart and lets us taste eternal life from the eternal fountain of life.

For this is the truth, the eternal, life-giving truth, which the heart appropriates only by the Spirit of truth: that Jesus has gone forth from the Father and come into the world; and he has left the world again and gone to the Father; he came with life from God to a race of men who had lost the life of God; he returned to the Father in order to lead those whom he saved into eternal glory. The regained, restored, and given back image of God — this is God’s truth in Jesus Christ and in us through the Spirit of Jesus.

Therefore the Spirit of truth is also the Spirit of life. “I live, and you will live,” says Jesus. The same Spirit who is the Spirit of truth, and whom the world does not know because it lives in the deceit of sin, the same Spirit is also the Spirit of life. For if the truth is the image of God renewed in our hearts, then this is life, that we become children of God when the Father’s image is impressed upon our soul.

Love Jesus! 

And you are partaker of his Spirit; the truth will shine into your heart, and the life will penetrate your whole being. The old has passed away, and all has become new by the Spirit of truth.

On that day you will experience that Jesus is in the Father, and you in him and he in you. Then you have fellowship in the light with the Father and the Son by the Holy Spirit. Your soul, which has gone forth from God, has found its way back to him; and you are blessed in the fellowship of love with God.

It is your Pentecost day, O soul. It is as a gentle and blessed spring after a long winter. It is as the clear morning after black night. It is as lovely dew upon dry field, as refreshing rain after drought. Blessed is he whose life has had such a Pentecost day.

It is time, friends, for us to awake and arise from sleep, that it may be evident that we are children of the light, and that the night is now past.

This is Pentecost—the hour when God’s church was born on earth. Shall it be for us only a dead memory of long-vanished greatness? O no, let us awake as is right, and give the Spirit room in our hearts, each one and all together, and then God’s church will again take form among us, the Body of Jesus Christ again receive its growth among us, the Bride of Jesus again be adorned for her heavenly Bridegroom.

Awake, awake, put on your strength, O Zion; put on your beautiful garments, O Jerusalem, the holy city. Shake yourself from the dust, arise, sit upon your throne, O Jerusalem; loose yourself from the bonds of your neck, you captive daughter of Zion.

Long enough has the glory of God’s church been dimmed among us; step forth again in the strength of the Spirit and new life from God. Then there will be a new Pentecost on earth, and your light will shine to the ends of the world.

