

\section{Second Day of Pentecost: The Firm Foundation of Faith}

\begin{quote}

John 12:44–50. Jesus cried out: Whoever believes in me does not believe only in me, but in him who sent me; and whoever sees me sees him who sent me. I have come as a light into the world, that whoever believes in me will not remain in darkness. And if anyone hears my words and does not believe, I do not judge him; for I did not come to judge the world, but to save the world. Whoever rejects me and does not receive my words has his judge already: the word I have spoken will judge him on the last day. For I have not spoken on my own authority; but the Father who sent me has given me command what I am to say and speak; and I know that his commandment is eternal life. So what I speak, I speak as the Father has told me.

\end{quote}

\bigskip

This word, which Jesus cried out among the Jews, is of particular and solemn significance. It concerns the firm foundation upon which faith builds when it builds upon Jesus. It therefore also concerns the dreadful responsibility that follows upon despising and rejecting him.

A human soul requires firm ground on which to build. It is immortal, and if it is to find a lasting foundation on which to build, it cannot be something that perishes. Take all the best and strongest that the world has to offer, and it will soon show itself, under honest testing, that however good it may seem, yet it is not something upon which one may rely for eternity. The world passes away with its desires; and either the world leaves us, or we leave the world, and in either case our trust is gone and our confidence in vain. Blessed therefore is he who can say with the Psalmist: Though my flesh and my heart fail, God is the rock of my heart and my portion forever.

Thus the rock of the heart is the Lord himself. Nothing less is required. Therefore Jesus testifies this so powerfully, that whoever believes in him does not believe in a man, but in God who sent him. Jesus is sent into the world to reveal the Father and the Father’s eternal counsel of salvation for souls. Sin had made a separation between God and men, between the Father and his children. And sin blinded the eyes with the false glory of the world and closed them to the true and real beauty of heaven and of the Father’s home; and instead of seeking the eternal and imperishable things, the hearts of men are drawn to the sinful and the sensual. Then God sent his Son, that he should be the Light of the world. What was hidden from men by their sin and by the threats of their evil conscience — God’s eternal love, the Father’s compassionate grace — this was to be revealed through the Son. And in the midst of the restless, deceitful, faithless, and disappointing world, Jesus should be the firm rock and the true light.

It is God’s own eternal heart that is revealed in Jesus. Now we may look into the Father’s love, if only we turn with heart and eye toward the eternal Light. So long as we walk turned away from God, with our back toward the Light, we remain in darkness; and without peace and without faith we walk uncertain and insecure forward toward a dark and hopeless night of eternity. The rock is there, but we do not know it. The Light is there, but we do not see it. But the day we turn, and the Light from Jesus’ loving eye falls into our heart, and we come to see that the Father loves us, then everything becomes different.

For Jesus has not come that we should remain in sin and in darkness, but that we should be saved from them. If he is the rock, then we must build on him; if he is the light, then we must walk in its brightness. It does you no good to know that the rock of salvation is there quite near you if you never set foot on it. It does you no good that the light shines around you, if you close your eyes and deliberately walk in blindness.

If the eternal rock is here in Jesus and in his eternal word — the word of God’s love sealed with Jesus’ own blood and death and resurrection from the dead — then this is what matters for us: that we build upon this foundation and take hold of this salvation. The firm foundation of faith is neither our feelings nor our understanding nor our conversion nor our prayer; the firm foundation of faith is Jesus himself; for he who believes in him believes in God who sent him. Can there be anything firmer or more reliable? Impossible; he who has the Son has also the Father; and he who has the Son has eternal life in him.

With anything less the soul cannot be satisfied. For when the form of this world passes away, and when all that is visible and of the senses disappears for us in death, then this alone is our consolation: Though I walk through the valley of the shadow of death, I fear no evil; for you are with me; your rod and your staff, they comfort me.

Will you then, O soul, have firm ground for your faith? Give yourself wholly over to Jesus. He has come to save the world; he has come to save you. You are welcome as you are, however you may appear when you come. Jesus will save you; that is his unchangeable promise, and upon that you may come in full assurance. All who have ever come to him have experienced this, that he casts none out.

But do not soothe yourself with this, that Jesus has come to save the world, and therefore there is no danger. If you despise his word, which came to save you and which loves you with eternal love, then be certain that you have that which judges you. The rejected Savior and the rejected love will become your judgment and an eternally burning fire for your lost soul. God willed to save you, and you would not be saved — who then will be able to rescue you from destruction? Can your sin and the pride of your heart, or the devil, save you? On the day the Lord is revealed you will experience that the word which he has spoken, and which you have despised, that will judge you, and none will save you from its judgment.

For Jesus’ word is not from himself, but from the Father. Therefore the word is also eternal life. Jesus is both the Light that shines in the darkness and the Life that overcomes death. Have you room for him in your heart; have you faith in him in your soul? This is what matters. For as many as received him, to them he gave authority to become children of God and to have eternal life.

Build therefore upon the rock and believe in Jesus without wavering. Do not be satisfied that you have heard about the rock, but set your foot upon it without fear, and you will stand firm when all falls. Let death come, let the heart break, let hell threaten, let the world perish — if you stand upon the rock, then know that you are secure and will not be shaken; for the rock of the heart is the Lord himself. 

Amen.

