
\section{All Saints’ Day: Blessed and Unblessed}

\begin{quote}

Luke 6:20-26. And He lifted up His eyes upon His disciples and said: Blessed are you poor, for yours is the Kingdom of God. Blessed are you who now hunger, for you shall be filled. Blessed are you who now weep, for you shall laugh. Blessed are you when men hate you, and when they cast you out and revile you and cast out your name as evil for the Son of Man’s sake. Rejoice in that day, and leap for joy! For behold, your reward is great in Heaven. In the same way their fathers did to the prophets. But woe to you rich! For you have already received your consolation. Woe to you who are full! For you shall hunger. Woe to you who now laugh! For you shall mourn and weep. Woe to you when all men speak well of you! For in the same way their fathers did to the false prophets.

\end{quote}

\bigskip

Jesus’ sermon is altogether serious and severe, and yet blessed and full of comfort. These are words that can bring us to earnest self-examination, and that can lay us low in deep contrition; they are also words that can bring us sweet consolation on our pilgrim-wandering here below.

He sets as the condition of blessedness the opposite of what the world counts happiness; and He cries woe over those who possess happiness according to the judgment of the world.

The first thing needed in order to become happy is to lose everything. For poor is the one who has nothing. But to the one who has nothing, the Kingdom of God is given for nothing. And blessed is the one who receives the Kingdom of God. Everything that a human being has in the world is perishable; blessedness therefore is not to be found in it, for blessedness is eternal happiness, happiness that is never lost, a treasure which moth and rust cannot consume, which perishability cannot touch. Blessed therefore is the one who lets go of all earthly and perishable things and receives instead the eternal, heavenly, and imperishable things.

But this does not take place without suffering. As hunger is tormenting and painful, so the heart’s demand for blessedness is a dreadful torment so long as it has not been satisfied by God. If the world’s food has shown itself to be stones instead of bread, swine’s food instead of human nourishment, then the soul suffers unspeakably until it is filled with the familiar fare of the Father’s house. It will not do to be full of the things of the world, if you would enjoy the good things of the Lord’s house; they are for the hungry, who are no longer satisfied by the world.

It does not go without tears; for the mind that is content and glad in the bondage of perishability is shut up against heavenly joy. There is certainly also a worldly weeping, which is just as bad as the world’s laughter, a sorrow of the world, which works death. It is not that which Jesus praises. He means the soul’s weeping, which springs from its longing after God and His fellowship, when nothing in the world can give refreshment and comfort, the sorrow of the heart when the whole world has shown itself to be a disappointment and a deceit to the soul, the sorrow of the heart over this, that the happiness of life has been lost through sin—it is that sorrow which prepares the way for heavenly comfort.

Nor does it go without conflict. For when the world perceives that a soul no longer allows itself to be deceived by its false riches, by its swine’s food, by its merry and hollow laughter, then all its bitterness awakes. It feels itself rejected and set aside; it feels itself like a deceiver who has been caught in his deceit, like a hypocrite from whose face the mask has been torn away, and it rages in wrath because someone has dared to proclaim what it and its goods are, because someone has seen through its hollowness and emptiness and turned to other and better things and begun to ask after the Kingdom of God and the blessedness of Heaven. The one who rejects the world is rejected by the world. In contempt the world will turn its back upon him and make his name an object of mockery and scorn. In malice it will present him as a hypocrite, because its own falsehood has been laid bare.

But blessed, blessed is the one who, hungering and thirsting after truth, weeping and sighing after joy, has turned from the world and has found Jesus. For it must be so. There is no other way of becoming partaker of the heavenly, invisible, and eternal goods than the way which is Jesus Himself. He is the One who has come down from Heaven to bring heavenly life and joy and blessedness and refreshment to the poor, the hungry, the weeping, and the rejected. He has come and graciously invites all the weary and heavy-laden to find rest with Him.

And He shuts no one out; for He does not require that you bring Him anything. He demands nothing from you; He will give you everything, only come empty-handed, so that you may receive His gifts. But if hand and heart and mouth are full of the world and its joy and its food, how then can He give you heavenly joy and heavenly bread? He sets no condition for your salvation except this one alone, that in your shipwreck in the world you must let go of the straw which you still clutch in your hand, and take hold of His outstretched Savior’s hand. O let go your hold upon the world’s perishability, and then let yourself sink down into the faithful arms of mercy, and you are saved. Blessed, blessed! Leap for joy, for you lost that which drew your heart down into perdition, and you gained the Friend who alone is mighty to bear you into Heaven’s eternal joy. Then you gained the happiness which is hidden from the world, but revealed to those who love God.

But unblessed is the one who is rich and full and glad in the world and praised by men. For the more the world fills and satisfies the heart, the more worldly happiness smiles upon us and beguiles us with its allurements, the more we delight in the world and the world delights in us, the more dreadful will be the day on which it all vanishes. What then remains for the worldly mind and heart when death comes, when the world fades, when everything, yes, all of it, is gone, gone forever? When there is only the God whom you hate, and the Hell which you fear, and nothing, nothing else to all eternity? How dreadful to have one’s eyes opened to the long self-deception when it is too late! Too late! O hasten, before the last beam of the sun of grace vanishes into the night of death and Hell; hasten to let go of the enchanting, ravishing, seductive world, which deceives you out of your happiness and blessedness, and lay hold on the eternal life which is in the Lord Jesus Christ and in no other.

For the poor, the hungry, the weeping, and the rejected belong the Kingdom of God and the blessedness of God. It has room for all human hearts without exception, if only they will make room for the Kingdom of God. There is no one who cannot become poor, if he will let the Spirit convict him of sin. There is no one who cannot become hungry, if he will hold himself back from the world’s swine’s food. There is no heart in which there cannot arise tears and sorrow and lamentation, if it will see its perdition in the light of God’s Word. Grace for all who need grace; blessedness for all who will know their unblessedness. God’s fountain is full of water, if only you would come and drink and take the water of life freely. Then you would experience how blessed that is which no eye has seen and no ear has heard, that which God has prepared for those who love Him.

Deliver me from the world!
Hide me, ah, hide me in You!
The joy in the world
Is indeed no joy for me.

