
\section{Twenty-Second Sunday after Trinity: He who has ears to hear, let him hear}

\begin{quote}

Mark 4:21-25. And He said to them: Does the light come in that it should be set under the bushel or under the bench? Is it not rather that it should be set upon the lampstand? For nothing is hidden except that it should be made manifest, neither has anything come to pass in order to remain concealed, but that it should come to light. If anyone has ears to hear, let him hear! And He said to them: Take heed what you hear! With the measure you measure, it shall be measured to you, and to you who hear there shall be given yet more. For he who has, to him shall be given; and he who has not, from him shall be taken even that which he has.

\end{quote}

\bigskip

To hear as well as to speak are two common and therefore seemingly unremarkable functions of two natural human organs of sense. Nevertheless, both are of such importance that not only life and death in the purely natural sense often depend upon them, but eternal life and eternal death also rest upon the Word and the right hearing of it. For there is a mutual relation between the two.

Just as it is said of the Word that it is a two-edged sword, which reveals the secret counsels of the hearts, and that for some it becomes a savor of death unto death, for others a savor of life unto life, so it is said of hearing: "Faith comes by hearing, but hearing takes place through the Word of God" (Rom. 10:17). Therefore the Lord says in our text: "Take heed what you hear!" And in another place: "See therefore how you hear!" (Luke 8:18).

A person’s salvation truly depends upon what he hears and how he hears.

It is impossible for one who has hearing to avoid hearing what is spoken and uttered all around him. But it is not long before he begins to make a difference and to prefer hearing one thing rather than another.

The ancients said: "Tell me with whom you keep company, and I shall tell you who you are." One could just as well recast it thus: "Tell me what you delight to hear, and I shall tell you who you are."

Some like to go where they hear spiritless and frivolous talk, such as a dog would be ashamed of if it were written down and read aloud again. They do so for pastime, they say, and readily add some sort of amusement or game besides.

Others prefer a convivial circle where they may hear gossip and evil report and slander of their fellow men, and they drink in such speech as the sweetest draught.

Or they gather with those who let foul speech proceed out of their mouth together with mocking and frivolous jesting, immoral songs, oaths, and blasphemies, and their soul is permeated through and through with venom and poison from hell.

Some go much higher. They set value only upon hearing elevated subjects discussed: science, music, art, poetry, and all such things as require so costly an education that only a few can hope to gain entrance into that esteemed society. In the same proportion they find their pleasure and delight in it.

... all these different kinds of hearing have their own peculiar and decisive influence upon the condition of a person’s soul; but none of them brings peace. And however much anyone may comfort himself with being a Christian, though he has his delight in lending his ear to one or more of the above-mentioned ways of hearing, and thinks he can do so unharmed, he deceives himself. No one can withdraw himself from the influence of that to which he listens with delight and gladness; daily experience convinces us of it.

There is only one thing that truly does the soul good to hear; it is that which requires neither art nor learning to grasp, which need not hide itself "under bench or bushel" either; it is that which, as a light upon a lampstand and a city upon a hill, shines both for all the world and for each individual person with such heavenly power that it penetrates into the soul’s innermost depth and reveals all that is concealed: the most secret counsels and thoughts, the whole nature and abomination of sin; it is that which sounds now like a trumpet of judgment concerning the fire that is never quenched and the worm that never dies, now like a still, loving voice from heaven: "Though a mother forget her nursing child, I have not forgotten you!"—it is that by which the thorn-crowned and ascended Savior Himself is designated: the Word, the Word of God.

"Hear the Word of the Lord, you princes of Sodom," cries the Prophet; "hear My voice, and I will be your God, and you shall be My people." "Your words were found by me," says Jeremiah, "and I ate them, and Your words were to me joy and the gladness of my heart."

"Earth, earth, earth, hear the Word of the Lord!"
Friend, you have ears; do you hear the Word of God? Oh, then you are blessed, if you do thereafter, that is to say, if you hear rightly.
For, says the Lord, "See therefore how you hear."
The Word does not return empty. There, where the Gospel is heard, it works either life and blessedness or judgment and death; it is—as it is said of the Son of God—set for the fall and for the rising again of many in Israel, and for a sign that is spoken against. He who hears the Word and does thereafter builds his house upon a rock; but he who does not do thereafter sets his house upon sand, and its fall shall be great.

A great and holy responsibility therefore rests upon every one who hears the Word of God, and to whom this precious treasure of life is thus entrusted.

To Israel the Lord says: "From the day your fathers came out of the land of Egypt unto this day, I sent to you all My servants the prophets, daily, early, and continually. But they did not hear Me, and did not incline their ear; they hardened their neck, they acted worse than their fathers. Therefore I will cast you out from before My face, just as I cast out your brethren, all the seed of Ephraim."

Does not that same Word and that same judgment still sound to us?
"With the measure you measure, it shall be measured to you again, and to you who hear there shall be given yet more—but from him who has not there shall be taken even that which he has."

In this word of our text the Lord exhorts to carefulness in one’s dealing with the Word of God, and to keep nearer to hearing than a fool to offering sacrifice.

If with prayer and crying out for the Spirit’s blessing, with a heart hungering and thirsting, you go to hear or read the Word of God, so that you think you could gladly swallow the whole of it in order to satisfy your soul, then the Lord shall give you according to your sighing and your desire, "a good measure, pressed down, shaken together, and running over, into your lap." "In joy you shall go out and in peace you shall be led forth; the mountains and the hills shall break forth into shouts of joy before your face, and all the trees of the field shall clap their hands" (Isa. 55:12).

But if you go to the Word with indifference, with the proud thought that you yourself can master it, without hunger and thirst, for mere appearance’s sake or outward form, then your heart shall for each time become emptier and emptier, more spiritless and more spiritless, so that the words spring back from your soul like a ball cast against a wall, until you become hard as stone, and judgment comes upon you unawares as upon Israel. Then there is taken from you even that which you had.

And now, Friend, how often have you been in God’s house, how often have you listened to His Word about sin and grace, about death and life, how lovingly has the Lord not called to you, how gently and earnestly has not the Father drawn you again and again, and how do you hear? Is it with delight and longing, are the words to you sweeter than honey and the honeycomb, do you draw water with joy out of the springs of life, or are you sluggish and dull, is it only reluctantly or by habit that you receive the Savior’s message, is it a relief to you when the preaching of the Word is over, do you hasten with double eagerness to your newspaper, your neighbor’s talk, or to your books and speculations?

Brother and Sister! Do you remember that the Lord has spoken to you daily, early, and continually, that the hourglass is soon run out, that you bear responsibility for all the words God has spoken to you in vain—that you are in the process of becoming hardened, and that after that comes the judgment?

He who has ears to hear, let him hear!

